Analyzing the Effect of Surface in Tennis Grand Slams

### 2018-09-15 contest/surface-type/paper-tennis.Rmd
dreaded merge

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = TRUE)
library(deuce)
library(dplyr)
library(ggplot2)
library(forcats)
library(stringr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(grid)

## To change into library(courtsports)
devtools::load_all("../..")
```

@@

# Results {#sec:results}

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(cache = TRUE)
library(deuce)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(stringr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(grid)
library(MASS)
## To change into library(courtsports)
devtools::load_all("../..")

@@

## Examining individual players
```{r table}
data(gs_partial_players)

top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
matches_played_slam_year <- top_three %>% group_by(name, tournament, year) %>%
  summarize(matches_played = n())
matches_played_slam <- matches_played_slam_year %>% group_by(name, tournament) %>%
  summarize(matches_played = sum(matches_played))
tab <- spread(matches_played_slam, key = name, value = matches_played)
total_df <- data.frame(tournament = "Total", as.list(colSums(tab[,-1])))
colnames(total_df) <- colnames(tab)
tab <- rbind(tab, total_df)

knitr::kable(tab, format = "latex",
               row.names = FALSE, col.names = c("Tournament", "Nadal", "Federer", "Williams"),
               caption = "\\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.",
               booktabs = TRUE) %>% kable_styling(latex_options = "striped") %>%
    row_spec(5, bold = TRUE, background = "lightgray")
```

In Table \ref{tab:three-gs-counts}, we display the number of matches Nadal, Federer, and Williams have played from 2013-2017.  Over that time span, Nadal won 6 grand slams, Federer won 1, Williams won 8.  Despite this, Federer played more total matches in Nadal.  All three players were absent for exactly three slams during this time period due to external factors (Nadal: (1 AO, 0 FO, 1 Wim, 1 USO), Federer: (0 AO, 2 FO, 0 Wim, 1 USO), Serena: (0 AO, 1 FO, 1 Wim, 1 USO)).  Not unexpectedly, Nadal has the most wins on clay (`r tab[2, 2]`), Federer has the most wins at Wimbledon (`r tab[4, 3]`) and Williams the most on hardcourt (`r tab[1, 4]` at AO and `r tab[4, 4]` at USO).  Despite having played more matches than Nadal, Federer only has 1 grand slam to show compared to Nadal's six, which indicates that Federer made it deeper into tournaments on average but had difficulties winning the championships.  For the WTA, Williams had her second most successful five-year span at the grand slams over this time period, winning 8.

We fit individual models for these three individuals (subsetting the data to their matches only), estimating the percent of points won in a match using stepwise regression.  The lower model is the  percent of points regressed on the opponent rank and indicator variables for court type with the FO as the reference variable.  The upper model is the lower model with the additional variables of winner to unforced error ratio, average serve speed, percent aces, percent break points, percent net points won, and their interaction effects with court type.  We do not display the full best fit models here, but they are available online.

ADD CIS!!

## Nadal

```{r nadal-model}

step_model_n <- model_individual(player_name = "Rafael Nadal", data = gs_partial_players)

std_res_n <- rstandard(step_model_n)

gn <- ggplot(lm(step_model_n$formula, data = step_model_n$data)) + geom_hline(yintercept = 0, col = "red") +
  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
  labs(title = "Nadal",
       x = "Predicted Percent of Points Won", y = "Standardized Residuals")

coefs_n <- coefficients(step_model_n)
```

Looking at Nadal individually, the model with the lowest AIC has positive coefficients opponent rank, percent of aces, and percent of break points won.  It has negative coefficients for AO, USO, and Wim. compared to FO and for percent of net points won.  There are no interaction effects in this model.

The standardized residuals plot shown in Figure \ref{fig:individ-resids} (along with the other linear regression diagnostics plots not shown here) shows that the model is a good fit for predicting the percent of points won.  For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_n[2]), 2)`$ less at the AO and $`r signif(abs(coefs_n[4]), 2)`$ less at Wim compared to the FO; to increase by $`r signif(abs(coefs_n[5]), 2)`$ for a one unit increase in opponent rank; a `r .01 * signif(abs(coefs_n[7]), 2)`\% increase for a .01 increase in percent of break points won; and to decrease by $`r .01 * signif(abs(coefs_n[8]), 2)`$ for a .01 increase in percent of net points won.  As such, we see clear evidence Nadal performs better at the FO compared to the other grand slams.

## Federer
```{r federer-model}

step_model_f <- model_individual(player_name = "Roger Federer", data = gs_partial_players, ref = "Wimbledon")

std_res_f <- rstandard(step_model_f)

gf <- ggplot(lm(step_model_f$formula, data = step_model_f$data)) + geom_hline(yintercept = 0, col = "red") +
  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
  labs(title = "Federer",
       x = "Predicted Percent of Points Won", y = "Standardized Residuals")

coefs_f <- coefficients(step_model_f)
```

For Federer, the model with the lowest AIC has negative coefficients AO, FO, and USO. vs Wim., and W/UE,  It has positive coefficients for opponent rank, percent of break points won, and all the interaction terms: court and W/UE.

The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but almost seem to see two clusters of residuals splitting with predicted point percentage of about 0.63.  For the significant variabless ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to be $`r signif(abs(coefs_f[2] + coefs_f[8]), 2)`$ less at AO compared to the Wim.; $`r signif(abs(coefs_f[3] + coefs_f[9]), 2)`$ less at FO compared to the Wim.; $`r signif(abs(coefs_f[4] + coefs_f[9]), 2)`$\% less at USO compared to the Wim.;  to decrease by $`r signif(abs(coefs_f[6] + coefs_f[8]) , 2)`$ for a one unit increase in W/UE at AO, $`r signif(abs(coefs_f[6] + coefs_f[9]) , 2)`$ for a one unit increase in W/UE at FO, and $`r signif(abs(coefs_f[6] + coefs_f[10]) , 2)`$ for a one unit increase in W/UE at USO.  It is clear that Federer performs best at Wimbledon.

## Williams
```{r williams-model}

step_model_w <- model_individual(player_name = "Serena Williams", data = gs_partial_players,
                                 start_lower = TRUE, ref = "Australian Open")

std_res_w <- rstandard(step_model_w)

gw <- ggplot(lm(step_model_w$formula, data = step_model_w$data)) + geom_hline(yintercept = 0, col = "red") +
  geom_point(aes(x=.fitted, y = .stdresid)) + my_theme + ylim(-3, 3) +
  labs(title = "Williams",
       x = "Predicted Percent of Points Won", y = "Standardized Residuals")

coefs_w <- coefficients(step_model_w)
```

```{r resid plots,  fig.height=4, fig.width=12, fig.cap = "\\label{fig:individ-resids}Standardized residuals vs percent of points won  best fit model from forward-backwards stepwise regression for Nadal's, Federer's, and William's, respectively.}"}

grid.arrange(gn, gf, gw, ncol = 3, 
              top = textGrob("Standardized Residuals vs. Predicted % Points Won",
                            gp=gpar(fontsize=20, fontfamily="serif")))

For Williams, the model with the lowest AIC is the largest model of the three selected 22 coefficients.  Since Williams only has 59 observations, we believe this model is overfitting and so caution against inference with this model.  It should also be noted that no women's point by point data was recorded for 2015 so that year is excluded from this model.  The coefficients are court, opponent rank, W/UE, average serve speed, percent aces, percent break points won, percent net points won, and interaction effects with court and W/UE, court and average service speed, court and percent aces, and court and break points won.

The standardized residuals plot shown in Figure \ref{fig:individ-resids}  shows that the model is a fair fit for predicting the percent of points won, but we tend to overestimate the percent of points won in comparison with Nadal.   For the significant variables ($\alpha$ = 95\%), while adjusting for the other variables, we expect percent of points won to increase by $`r signif(abs(coefs_w[5]), 2)`$ for a 1 unit increase in oponent rank; to increase by $`r signif(abs(coefs_w[6] + coefs_w[11]) , 2)`$ for a one unit increase in W/UE at FO, to increase by  $`r signif(abs(coefs_w[6] + coefs_w[12]) , 2)`$for a one unit increase in W/UE at USO, and to increase by  $`r signif(abs(coefs_w[6] + coefs_w[13]) , 2)`$ for a one unit increase in W/UE at Wim. compared to AO; to increase by  $`r signif(.01 * abs(coefs_w[8] + coefs_w[17]) , 2)`$ for a .01 increase in percent of aces at the French Open; and to increase by $`r signif(.01 * abs(coefs_w[9] + coefs_w[20]) , 2)`$ for a .01 increase in percent of break points won at the French Open.

For this, we do not see that Williams is dominant on the hard courts compared to the other courts.  Rather, she has good results on all the courts.  Her significant interaction effects show that if she focuses on acing her opponent and capitalizing on breakpoints, then she has a better chance at winning more points than at the Australian Open.

### 2018-09-15 contest/surface-type/paper-tennis.tex
dreaded merge

  \title{Analyzing the Effect of Surface in Tennis Grand Slams}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{September 12, 2018}

@@

We use models that take into account both individual and group effects
such as in the Gaussian-process player production basketball model or
predicting individual soccer performance (Page, Barney, and McGuire
2013; Egidi and Gabry 2018). Both of those models had success using
hierarhcical Bayesian models, which we employ in our own models. More
specifically, we model the players' expected points in a match based on
the player's own characteristics, the court/tournament effects, and the
oponent's ranking.

@@

The data consists of 5080 matches split evenly over the four grand slams
and the two leagues: ATP (men's) and WTA (women's). Each match has 80
attributes, many of which are redundant. We focus on the following
attributes for both the winner and loser of the match: games won, points
won, retirement, break points faced, break points saved, aces, country
of origin, and player attributes. Additionally, we take into account the
number of sets in a match, the surface type, minutes played, and round
of the tournament. A subset of the data is shown in Table
\ref{tab:data}.

@@

@@

\hypertarget{sec:results}{%
\section{Results}\label{sec:results}}

>>>>

  \title{Analyzing the Effect of Surface in Tennis Grand Slams}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{September 15, 2018}

@@

We use models that take into account both individual and group effects
such as in the Gaussian-process player production basketball model or
predicting individual soccer performance (Page, Barney, and McGuire
2013; Egidi and Gabry 2018). Both of those models had success using
hierarchical Bayesian models, which we employ in our own models. More
specifically, we model the players' expected points in a match based on
the player's own characteristics, the court/tournament effects, and the
opponent's ranking.

@@

The data consists of 5080 matches split evenly over the four grand slams
and the two leagues: ATP (men's) and WTA (women's). Each match has 80
attributes, many of which are redundant. We focus on the following
attributes for both the winner and loser of the match: games won, points
won, retirement, break points faced, break points saved, aces, country
of origin, and player attributes. Additionally, we take into account the
number of sets in a match, the surface type, and round of the
tournament. A subset of the data is shown in Table \ref{tab:data}.

@@

\hypertarget{examining-the-distribution-of-points-earned}{%
\subsubsection{Examining the distribution of points
earned}\label{examining-the-distribution-of-points-earned}}

We first examine the distribution of points earned per match. We find
that the distribution is similar across tournament, with Wimbledon
differing slightly from the other grand slams. As expected, there are
more points earned in the WTA than the ATP due to the differing numbers
of games played. Also unsurprisingly, the winners of the match tended to
earn more points than the losers.

\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-1-1.pdf}

\hypertarget{home-court-advantage}{%
\subsubsection{Home court advantage}\label{home-court-advantage}}

It is commonly thought that there is a home court advantage in grand
slam games (SOURCE). In our data we find this to be true (i.e.~French
players win the French open more than French players win other slams).
But, we also know that the home team is given preference for wild card
bids (SOURCE) so potentially citizens of a particular country play in
``their'' tournament more often than they play in other tournaments. We
also find this to be true in our data for France, The United States, and
Australia (i.e.~the proportion of French players in the French Open is
greater than the proportion of French players in other slams).

Therefore, we want to see how the proportion of wins for the home
country changes across the different tournaments. If there was really a
home court advantage, the proportion of French wins each round would be
higher at the French Open than the Australian Open, the US Open, and
Wimbledon. The same would be true for Australia and the US. But, we see
that this isn't the case. After accounting for the number of players
from each country, we don't find a home court advantage in the grand
slam.

\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-2-1.pdf}

\hypertarget{spaniards-on-clay}{%
\subsection{Spaniards on clay}\label{spaniards-on-clay}}

It is also commonly thought that Spaniards play better on clay. We are
interested in whether Spaniards win the French Open more than they win
in other tournaments. It does appear that Spaniards are winning the
French Open more than they are winning other tournaments. But, this
result is not significant. In addition, the ranking of Spaniards in the
French Open is, on average, higher than other tournaments, which may
help explain this common misconception, although this result is not
significant either.

\includegraphics{paper-tennis_files/figure-latex/unnamed-chunk-3-1.pdf}

@@

\hypertarget{examining-individual-players}{%
\subsection{Examining individual
players}\label{examining-individual-players}}

\caption{\label{tab:table}\label{tab:three-gs-counts}Number of matches played for Nadal, Federer, and Williams from 2013-2017 at each of the grand slams.}
\centering
\begin{tabular}[t]{lrrr}
\hiderowcolors
\toprule
Tournament & Nadal & Federer & Williams\\
\midrule
\showrowcolors
Australian Open & 20 & 28 & 30\\
French Open & 29 & 14 & 23\\
US Open & 21 & 22 & 26\\
Wimbledon & 11 & 29 & 21\\
\rowcolor{lightgray}  \textbf{Total} & \textbf{81} & \textbf{93} & \textbf{100}\\
\bottomrule
\end{tabular}
\end{table}
\rowcolors{2}{white}{white}

In Table \ref{tab:three-gs-counts}, we display the number of matches
Nadal, Federer, and Williams have played from 2013-2017. Over that time
span, Nadal won 6 grand slams, Federer won 1, Williams won 8. Despite
this, Federer played more total matches in Nadal. All three players were
absent for exactly three slams during this time period due to external
factors (Nadal: (1 AO, 0 FO, 1 Wim, 1 USO), Federer: (0 AO, 2 FO, 0 Wim,
1 USO), Serena: (0 AO, 1 FO, 1 Wim, 1 USO)). Not unexpectedly, Nadal has
the most wins on clay (29), Federer has the most wins at Wimbledon (29)
and Williams the most on hardcourt (30 at AO and 21 at USO). Despite
having played more matches than Nadal, Federer only has 1 grand slam to
show compared to Nadal's six, which indicates that Federer made it
deeper into tournaments on average but had difficulties winning the
championships. For the WTA, Williams had her second most successful
five-year span at the grand slams over this time period, winning 8.

We fit individual models for these three individuals (subsetting the
data to their matches only), estimating the percent of points won in a
match using stepwise regression. The lower model is the percent of
points regressed on the opponent rank and indicator variables for court
type with the FO as the reference variable. The upper model is the lower
model with the additional variables of winner to unforced error ratio,
average serve speed, percent aces, percent break points, percent net
points won, and their interaction effects with court type. We do not
display the full best fit models here, but they are available online.

ADD CIS!!

\hypertarget{nadal}{%
\subsection{Nadal}\label{nadal}}

Looking at Nadal individually, the model with the lowest AIC has
positive coefficients opponent rank, percent of aces, and percent of
break points won. It has negative coefficients for AO, USO, and Wim.
compared to FO and for percent of net points won. There are no
interaction effects in this model.

The standardized residuals plot shown in Figure \ref{fig:individ-resids}
(along with the other linear regression diagnostics plots not shown
here) shows that the model is a good fit for predicting the percent of
points won. For the significant variables (\(\alpha\) = 95\%), while
adjusting for the other variables, we expect percent of points won to be
\(0.034\) less at the AO and \(0.046\) less at Wim compared to the FO;
to increase by \(1.5\times 10^{-4}\) for a one unit increase in opponent
rank; a 0.0012\% increase for a .01 increase in percent of break points
won; and to decrease by \(8.8\times 10^{-4}\) for a .01 increase in
percent of net points won. As such, we see clear evidence Nadal performs
better at the FO compared to the other grand slams.

\hypertarget{federer}{%
\subsection{Federer}\label{federer}}

For Federer, the model with the lowest AIC has negative coefficients AO,
FO, and USO. vs Wim., and W/UE, It has positive coefficients for
opponent rank, percent of break points won, and all the interaction
terms: court and W/UE.

The standardized residuals plot shown in Figure \ref{fig:individ-resids}
shows that the model is a fair fit for predicting the percent of points
won, but almost seem to see two clusters of residuals splitting with
predicted point percentage of about 0.63. For the significant variabless
(\(\alpha\) = 95\%), while adjusting for the other variables, we expect
percent of points won to be \(0.04\) less at AO compared to the Wim.;
\(0.058\) less at FO compared to the Wim.; \(0.007\)\% less at USO
compared to the Wim.; to decrease by \(0.022\) for a one unit increase
in W/UE at AO, \(0.073\) for a one unit increase in W/UE at FO, and
\(0.041\) for a one unit increase in W/UE at USO. It is clear that
Federer performs best at Wimbledon.

\hypertarget{williams}{%
\subsection{Williams}\label{williams}}

\begin{figure}
\centering
\includegraphics{paper-tennis_files/figure-latex/resid plots-1.pdf}
\caption{\label{fig:individ-resids}Standardized residuals vs percent of
points won best fit model from forward-backwards stepwise regression for
Nadal's, Federer's, and William's, respectively.\}}
\end{figure}

For Williams, the model with the lowest AIC is the largest model of the
three selected 22 coefficients. Since Williams only has 59 observations,
we believe this model is overfitting and so caution against inference
with this model. It should also be noted that no women's point by point
data was recorded for 2015 so that year is excluded from this model. The
coefficients are court, opponent rank, W/UE, average serve speed,
percent aces, percent break points won, percent net points won, and
interaction effects with court and W/UE, court and average service
speed, court and percent aces, and court and break points won.

The standardized residuals plot shown in Figure \ref{fig:individ-resids}
shows that the model is a fair fit for predicting the percent of points
won, but we tend to overestimate the percent of points won in comparison
with Nadal. For the significant variables (\(\alpha\) = 95\%), while
adjusting for the other variables, we expect percent of points won to
increase by \(4.8\times 10^{-4}\) for a 1 unit increase in oponent rank;
to increase by \(0.034\) for a one unit increase in W/UE at FO, to
increase by \(0.092\)for a one unit increase in W/UE at USO, and to
increase by \(0.0038\) for a one unit increase in W/UE at Wim. compared
to AO; to increase by \(2\times 10^{-4}\) for a .01 increase in percent
of aces at the French Open; and to increase by \(0.0024\) for a .01
increase in percent of break points won at the French Open.

For this, we do not see that Williams is dominant on the hard courts
compared to the other courts. Rather, she has good results on all the
courts. Her significant interaction effects show that if she focuses on
acing her opponent and capitalizing on breakpoints, then she has a
better chance at winning more points than at the Australian Open.
